### 2017-06-03 paper/sexSelectionSpacing-ver1.tex
Change reference/name for descriptive stat file

\begin{figure}[htp]%
\centering
\includegraphics[width=0.9\textwidth]{sex_ratio_marriage}
\caption{Ratio of boys in ``first'' births}
\label{fig:sexRatioMarriage}
\end{figure}

@@

\begin{figure}[htpb]
\centering
\caption*{Urban}
\setcounter{subfigure}{-2}
\subfloat[Spell 2 - 1 boy (N=3,969)]{
    \begin{minipage}{0.30\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell2_g3_high_r2_pc}}\\
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell2_g3_high_r2_s}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\setcounter{subfigure}{-1}
\subfloat[Spell 3 - 1 boy/1 girl (N=2,357)]{
    \begin{minipage}{0.30\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r4_pc}}\\
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r4_s}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\setcounter{subfigure}{0}
\subfloat[Spell 3 - 2 boys (N=1,098)]{
    \begin{minipage}{0.30\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r2_pc}}\\
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r2_s}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\caption*{Rural}
\setcounter{subfigure}{1}
\subfloat[Spell 2 - 1 boy (N=3,044)]{
    \begin{minipage}{0.30\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell2_g3_high_r1_pc}}\\
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell2_g3_high_r1_s}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\setcounter{subfigure}{2}
\subfloat[Spell 3 - 1 boy/1 girl (N=1,819)]{
    \begin{minipage}{0.30\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r3_pc}}\\
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r3_s}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\setcounter{subfigure}{3}
\subfloat[Spell 3 - 2 boys (N=790)]{
    \begin{minipage}{0.30\textwidth}
        \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
        \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r1_pc}}\\
        \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r1_s}}
        \captionsetup[subfigure]{labelformat=parens}
    \end{minipage}
}
\caption{Predicted probability of having a boy and probability of
no birth by quarter (3 month period) for women with 8 or more years
of education for 1995--2006. 
N is number of women in the relevant group in the underlying samples.
}
\label{fig:boys_latest}
\end{figure}

@@

\subsection{Predicted Lifetime Number of Children and Abortions}

One of the main advantages of the method proposed is that we can predict what 
completed fertility, total number of abortions, and the sex ratios will be for 
women in the samples at the end of childbearing. 
This cannot be done with the simple model because it does not predict fertility
progression, does not take into account censoring, and cannot capture if parents
change their use of sex selection within a spell.

The predictions are important for three reasons.
First, they allows us to quantify whether overall use of sex selection is 
intensifying or easing over time by examining how the total number of abortions 
women will have over their childbearing, and the resulting final sex ratios, change.
This is a more accurate representation than what we get from sex ratios 
for individual parities.
As the theory section shows, it is possible to get no change---or even a 
decline---in use of sex selection for an individual parity, at the same time 
as the overall use of sex selection is increasing.
Second, completed fertility and sex ratios are what matters when examining the future 
impacts of sex selection in areas such as the chance of finding a spouse, savings 
behavior, and the potential problems associated with a surplus of bachelors 
\citep{lancaster02,Ding2009,Wei2009,Edlund2013}.
Finally, as I show in the online Appendix and discuss above, relying only on observed 
sex ratios of recorded births is problematic because these are based on selection and 
the observed sex ratio may be biased predictors of final sex ratio when people change 
their use of sex selection during a spell.

Table \ref{tab:predicted} shows predicted birth progression rates, number of abortions, 
sex ratios, and completed fertility for the three samples of women with 8 or more years 
of education:  
the 11,271 women who were married in the the 1972-1984 period,
the 19,072 women who were married in the 1985--1994 period,
and the 16,620 women who were married in the 1995--2006 period.%
\footnote{
These are the women used to estimate the results for the spell from marriage to first 
birth above.
For more information see Table \ref{tab:des_stat1}.
}
Predictions for each period assumes that behavior remains constant and that all spells 
a woman goes through are within the same period. 
Each woman is assumed to follow the fertility and sex selection behavior for 
each spell up to and including the fourth spell using the period's estimates and her 
characteristics. 
All results are for women predicted to have one or more births.%
\footnote{
This means that fertility predictions are larger than standard estimates, which include
women with no children.
Working in the opposite direction is that number of children is restricted to be at most 
four.
}

The predictions are done as follows.
Starting with the first spell I calculate each woman's probabilities of having a boy,
$P(b_t)$, a girl, $P(g_t)$, or no birth, $1-P(b_t)-P(g_t)$, in each quarter after marriage
based on the period results for 
equations (\ref{eq:probability_boy}) and (\ref{eq:probability_girl}) using her individual 
values for $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$.
She then is assigned one of these outcomes---boy, girl, or no birth---based on random 
draws from a uniform distribution.
The first quarter she has a draw that results in a birth she leaves the spell
and the quarter reached is used to calculate her starting age for the second spell.
If she has no draw that result in a birth by the end of the spell her childbearing 
stops and she does not enter the next spell.
If she does have birth, the process is repeated for the second through fourth spells.
Randomization is necessary because each spell's results, except for the first, depends on
the sex composition of previous children and the duration of the first spell. 
The results presented are averaged outcomes over all urban women and all rural women 
using 1,000 repetitions.

Birth progression rates are the percentage of women predicted to have a birth 
by the end of each spell.
Fertility is the average number of children born according to the randomization.
With the large number of replications this number is the same as using progression
rates at the end of the spell to calculate the average number of children.%
\footnote{
Given that all numbers are conditional on having at least one child, using progression
rates for, say, urban women from the 1985-1994 period leads to 
$1+1\times0.782+1\times0.782\times0.499+1\times0.782\times0.499\times0.448=2.3 $  
children.
}

For abortions I first calculate each quarter's probability of abortions using
$\frac{100}{105}P(b_t)-P(g_t)$,
where $\frac{100}{105}P(b_t)$ is the probability of having a girl that should have 
prevailed in the absence of sex selection (0.952 girls born per boy).
This probability is multiplied with the probability of not having had a birth yet at 
the beginning of the quarter.
Abortions for each spell are then found by summing over the entire spell.%
\footnote{
One way to see how this works is to imagine that we start the spell with 100 women.
After the first quarter we are left with $N_2 = (1-P(b_1)-P(g_1)) \times 100$, where
$N_2$ is the number of women who start the second quarter. 
After the second quarter we then have $N_3 = (1-P(b_2)-P(g_2)) \times N_2$, etc.

Say that 65.88 women of the original 100 still have not had a birth yet by the start
of quarter 6 and that the predicted probability of having a son in quarter 6 is 0.0931 and 
the probability of having a girl is 0.0566 (the probability of not having a birth in 
quarter 6 is 0.8503).
For this quarter the abortion probability is then 
$\frac{100}{105}\times 0.0931 - 0.0566 = 0.0321$.
In other words, the probability of having a girl should have been 
$\frac{100}{105}\times 0.0931 = 0.0887$, but only a probability of $0.0566$ was 
predicted.
Hence, in quarter 6 the predicted number of sons born is $ 0.0931 \times 65.88 = 6.13$, 
the predicted number of daughters born is $ 0.0566 \times 65.88 = 3.73$, and 
$0.0321 \times 65.88 = 2.11$ abortions must have taken place to reach these births.
}
To account for random variation, the abortion rate is the sum over both positive 
(more boys than the natural rate) and negative (more girls than the natural
rate) values of the abortion calculation within the spell.
Finally, the percentage of children born that are boys is based on the randomization of
births described above.%
\footnote{
This number is the predicted sex ratio at birth and differ from actual observed sex 
ratios when the children are older because of mortality.
On one hand, the anticipated or experienced death of a son increases the likelihood
of sex selection even if parents already had a son, which would make the later observed 
sex ratio lower than what is predicted here. 
On the other hand, mortality risk is higher for girls than for boy, which would make
the sex ratio more male dominated.
}

% Tables of predictions 
\input{../tables/predictedFertility.tex}

The results show a substantial reduction in fertility over time.
Predicted completed fertility went from 2.7 for urban women married in  
the 1972--1984 period to just 2 for urban women married in the 1995--2006 period.
Hence, urban fertility is now below replacement, especially considering that
these predictions are conditional on having at least one child.
Rural women showed a similar decline in completed fertility, going from 3.2 to 2.6.
Corresponding to the decline in overall fertility, there were also substantial
reductions in the individual parity progression rates.
For example, urban women with 3 children went from having a 57\% chance of having a 
4th child for the 1972--1984 cohort to a 33\% chance for the 1995--2006 cohort---and 
they had a substantially lower probability of even making it to 3 children
in the first place.

Associated with the decline in completed fertility is an increase in sex selection.
For urban women the predicted number of abortions per 100 women by the end of
childbearing have gone up by over 17 percent between the last two cohorts (8.0 to 9.4) 
and the number of abortions per 100 births have gone up by over 35 percent (3.4 to 4.6).
For rural women, abortions per 100 women increased almost 55 percent (6.8 to 10.5), 
while abortions per 100 births increased over 70 percent (2.4 to 4.1).%
\footnote{
Unfortunately there are no solid official estimates to compare these numbers to.
The total number of abortions per year, not just those based on son preference, vary 
between the official count of 600,000--700,000 
and an estimate of 6.4 million from the 2002 Abortion Assessment Project-India.
The 6.4 million abortions per year may even be an underestimate \citep{Stillman2014}.
NFHS does ask about abortions in all three surveys, although the questions change 
substantially between surveys.
However, the PNDT act prohibits using prenatal diagnostic tests to determine a fetus' sex
and revealing a fetus' sex to parents.
Hence, respondents have a strong incentive not to disclose abortions motivated by son 
preference. 
In addition, abortion are considered a taboo subject and many women are unwilling or 
uncomfortable reporting their abortions to government sponsored enumerators 
\citep{Rossier2003,Stillman2014}.
}
The combined effects of the declines in fertility and increased use of sex selection
are that the percentage boys increases from 53.1 to 53.7 percent for urban
women, and from 52.8 to 53.4 percent for rural women.

Of particular note is the high number of abortions among rural women for the 
1985--1994 period.
Educated rural women obviously had few constraints on access to 
prenatal sex determination even when the techniques were relatively new.
The higher abortion rate per woman for rural women than urban women likely comes from
a combination of higher likelihood of having an additional child and changes in use of 
sex selection within spell as shown in the graphs above.

The results from theoretical model showed that it is possible for the use of sex 
selection to fall for an individual parity, even as overall use of sex selection 
increases. 
This is the case here, where the use of sex selection fell for the third spell 
from 9.8 per 100 births to 7.8 for urban women between the last two periods.
As already discussed, total use of sex selection went up between the last two
periods, so the reduction for the third spell is not an indication of reduced 
use of sex selection overall, but simply the result of fewer women making it to a 
third birth and those who did were more likely to already have a son---for 
the second spell the number of abortions per 100 births for urban women 
went from 4.7 to 8.0, a 70 percent increase.%
\footnote{
The amount of sex selection in the fourth spell reinforces this point:
abortions per 100 births went up almost tenfold.
Only 26 percent of urban women with at least one child make it to the beginning 
of the fourth spell in the last period, and less than 10 of all women have a fourth birth, 
but for those who do make it to the fourth spell the pressure to ensure a son is large.
}
This underscores why focusing exclusively on individual parities can be misleading.

The results for rural women illustrate the two different strategies for achieving one
son: either have many children or have few children and use sex selection.
For the 1986--1994 period there is no use of sex selection on fourth birth, which fits
with 62 percent of women end up with three or more children and a full 62 percent of women with 
three children have a fourth birth (recall that there is a 94 percent chance of having at 
least one son if you have four children).
The use of sex selection on second and third spells indicates that one group of rural
women follows the high fertility route, whereas the other group follows the pattern of
urban women and combine restricting fertility with use of sex selection.

Another way to see the spread of access to sex selection over time is to compare
urban women in the first period with rural women in the last period.
At 2.7 fertility is already relatively low for urban women in the 1972--1984 period and 
there is some evidence of sex selection for these women.
As discussed in Section \ref{sec:spell_def}, the higher than natural sex ratio for
urban well-educated women this early is likely a combination of early access to prenatal
sex determination and that some of the spells end in the 1985--1994 period when there 
was more widespread access.
Rural women in the 1995--2006 period had a fertility only slightly lower than that
of urban women in the 1972--1984 periode, but the number of abortions both per birth and 
per woman were twice as high.

For comparison Appendix Tables \ref{tab:predicted_low} and \ref{tab:predicted_med}
show predicted fertility, number of abortions, and sex ratio for the two other
education groups.
As expected, fertility for women with no education is substantially higher
and the decline over time much smaller than for the high education group.
Predicted fertility fell from 3.6 to 3.3 for urban women and from 3.7 to 3.5 for rural 
women between the 1972--1984 and the 1995--2006 periods.%
\footnote{
A caveat is that my estimates did not cover parities higher than 4, which could
bias downward the change in predicted fertility.
}
Consistent with the discussion above, there is evidence of sex selection for neither
urban nor rural women, with the sex ratio for all predicted births in the last two periods 
close to normal.%
\footnote{
The higher than normal sex ratios for the 1972--1984 period are most likely
remnants of recall error not removed by the sample restriction. 
}
Random variation makes it appears that some individual parities had a higher
than normal sex ratio, which, if studied in isolation, could be taken as evidence
of use of sex selection, and reinforces why supplementing analysis of individual 
parities with this analysis of the overall pattern is important.

Fertility for women with between 1 and 7 years of education was slightly lower than the 
no education group in the 1972--1984 period, but the decline over time was larger.
For the 1995--2006 period predicted completed fertility was 2.8 for urban women and
3.2 for rural women.
There is only limited evidence of sex selection for urban women in this education group.
The exception is for rural women in the last period where the overall sex ratio is
52.1\% boys, although given the high fertility for this group and the small sample
size for this education group it is unclear whether this simply reflects random variation.

The predictions discussed here provide further support for the idea that there are 
two alternative strategies to ensure the birth of a son, either have many children 
or use sex-selective abortions.
The best educated group mostly follow the low fertility, high use of sex selection
strategy, whereas the two lower education groups still have relatively high fertility
and little to no use of sex selection.
It is, however, clear that fertility is declining over time for all groups.
With lower fertility in both urban and rural areas, we are likely to see 
further increases in the use of sex-selective abortions among those who already
use sex selection, and among the groups who have not previously use sex selection
we will see beginning use of sex-selective abortions.

@@

% Descriptive statistics tables
\input{../tables/desStat.tex}

>>>>

% \begin{figure}[htp]%
% \centering
% \includegraphics[width=0.9\textwidth]{sex_ratio_marriage}
% \caption{Ratio of boys in ``first'' births}
% \label{fig:sexRatioMarriage}
% \end{figure}

@@

% \begin{figure}[htpb]
% \centering
% \caption*{Urban}
% \setcounter{subfigure}{-2}
% \subfloat[Spell 2 - 1 boy (N=3,969)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell2_g3_high_r2_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell2_g3_high_r2_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \setcounter{subfigure}{-1}
% \subfloat[Spell 3 - 1 boy/1 girl (N=2,357)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r4_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r4_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \setcounter{subfigure}{0}
% \subfloat[Spell 3 - 2 boys (N=1,098)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r2_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r2_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \caption*{Rural}
% \setcounter{subfigure}{1}
% \subfloat[Spell 2 - 1 boy (N=3,044)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell2_g3_high_r1_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell2_g3_high_r1_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \setcounter{subfigure}{2}
% \subfloat[Spell 3 - 1 boy/1 girl (N=1,819)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r3_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r3_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \setcounter{subfigure}{3}
% \subfloat[Spell 3 - 2 boys (N=790)]{
%     \begin{minipage}{0.30\textwidth}
%         \captionsetup[subfigure]{labelformat=empty,position=top,captionskip=-1pt,farskip=-0.5pt}
%         \subfloat[Prob. boy (\%)]{\includegraphics[width=\textwidth]{spell3_g3_high_r1_pc}}\\
%         \subfloat[Prob. no birth yet]{\includegraphics[width=\textwidth]{spell3_g3_high_r1_s}}
%         \captionsetup[subfigure]{labelformat=parens}
%     \end{minipage}
% }
% \caption{Predicted probability of having a boy and probability of
% no birth by quarter (3 month period) for women with 8 or more years
% of education for 1995--2006. 
% N is number of women in the relevant group in the underlying samples.
% }
% \label{fig:boys_latest}
% \end{figure}

@@

@@

% Descriptive statistics tables
\input{../tables/des_stat.tex}
